### 2018-03-08 paper/background.tex
Work on S2.3

\subsection{The venerable \Make: static dependencies and file modification times}

@@

\begin{minted}[frame=single]{makefile}
util.o: util.h util.c
    gcc -c util.c

@@

main.exe: util.o main.o
    gcc util.o main.o -o main.exe
\end{minted}

@@

\begin{minted}[frame=single]{makefile}
release.tar: main.exe README
    tar -cf release.tar main.exe README
\end{minted}

So far so good: if either \cmd{main.exe} or \cmd{README} is dirty, \Make will
recreate the release archive. But now imagine that we need to specify the files
that go into \cmd{release.tar} in a text file \cmd{release.txt}. The dependencies
of \cmd{release.tar} are no longer known statically: they \emph{depend on the
contents} of \cmd{release.txt}, which might not even exist before the build
starts, e.g. if it is produced by concatenating input files \cmd{release-bins.txt}
and \cmd{release-docs.txt}. Alas, makefiles cannot express such \emph{dynamic
dependencies}, requiring ad hoc workarounds such as \emph{build phases}, which
are problematic~\cite{hadrian}.

\subsection{\Excel: dynamic dependencies at the cost of minimality}

@@

\begin{minted}[frame=single]{bash}
A1 = 10
A2 = 20
A3 = 30
B1 = 2
C1 = SUM(A1:INDIRECT("A" & B1))
\end{minted}

The output \cmd{C1} here is analogous to the \cmd{release.tar} from the previous
example: one cannot statically determine the dependencies of the associated task.
To deal with this, \Excel takes a simple conservative approach: it marks all
cells with dynamic dependencies as dirty and therefore always recomputes them
during a rebuild. This guarantees that the build result is correct, but clearly
violates the minimality requirement: if the user modifies \cmd{A3}, \Excel will
recompute \cmd{C1} even though it does not transitively depend on \cmd{A3}.

Note that \Excel's approach to marking cells as dirty is a relatively minor
variation of the approach used by \Make: a cell is considered dirty if it was
modified since the previous build \emph{or if it is volatile}, where the notion
of volatility includes dynamic dependencies, as well as \emph{non-determinism}
that will be discussed in~\S\ref{sec-engineering}. We therefore put both \Make
and \Excel in the same category of build systems that associate a \emph{dirty bit}
with every file or cell. (To avoid the ``file or cell'' nuisance, we introduce the
abstract notion of \emph{keys} in~\S\ref{sec-vocabulary}.)

@@

\subsection{\Shake: dynamic dependencies with no remorse}

\begin{figure}[h]
\centerline{\includegraphics[scale=0.28]{fig/make-example-cutoff.pdf}}
\caption{An early cutoff example.\label{fig-cutoff}}
\end{figure}

@@

\subsection{\Bazel: a cloud build system}

@@

\subsection{Abstract vocabulary for build systems}\label{sec-vocabulary}

>>>>

\subsection{The venerable \Make: static dependencies and file modification times}
\label{sec-background-make}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{makefile}
util.o: util.h util.c
    gcc -c util.c

@@

main.exe: util.o main.o
    gcc util.o main.o -o main.exe
\end{minted}
\vspace{1mm}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{makefile}
release.tar: main.exe README
    tar -cf release.tar main.exe README
\end{minted}
\vspace{1mm}

\noindent
So far so good: if either \cmd{main.exe} or \cmd{README} is dirty, \Make will
recreate the release archive. But now imagine that we need to specify the files
that go into \cmd{release.tar} in a text file \cmd{release.txt}. The dependencies
of \cmd{release.tar} are no longer known statically: they \emph{depend on the
contents} of \cmd{release.txt}, which might not even exist before the build
starts, e.g. if it is produced by concatenating input files \cmd{release-bins.txt}
and \cmd{release-docs.txt}. Alas, makefiles cannot express such \emph{dynamic
dependencies}, requiring ad hoc workarounds such as \emph{build phases}, which
are problematic~\cite{hadrian}.

\subsection{\Excel: dynamic dependencies at the cost of minimality}
\label{sec-background-excel}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{bash}
A1 = 10
A2 = 20
A3 = 30
B1 = 2
C1 = SUM(A1:INDIRECT("A" & B1))
\end{minted}
\vspace{1mm}

\noindent
The output \cmd{C1} here is analogous to the \cmd{release.tar} from the previous
example: one cannot statically determine the dependencies of the associated task.
To deal with this, \Excel takes a simple conservative approach: it marks all
cells with dynamic dependencies as dirty and therefore always recomputes them
during a rebuild. This guarantees that the build result is correct, but clearly
violates the minimality requirement: if the user modifies \cmd{A3}, \Excel will
recompute \cmd{C1} even though it does not transitively depend on \cmd{A3}.

Note that \Excel's approach to marking cells as dirty is a relatively minor
variation of the approach used by \Make: a cell is considered dirty if it was
modified since the previous build \emph{or if it is volatile}, where the notion
of volatility includes dynamic dependencies, as well as \emph{non-determinism}
that will be discussed in~\S\ref{sec-engineering}. We therefore put both \Make
and \Excel in the same category of build systems that associate a \emph{dirty bit}
with every file or cell. (To avoid the ``file or cell'' nuisance, we introduce the
abstract notion of \emph{keys} in~\S\ref{sec-background-vocabulary}.)

@@

\subsection{\Shake: dynamic dependencies with no remorse}
\label{sec-background-shake}

\Shake was developed specifically to address the issue of dynamic
dependencies~\cite{mitchell2012shake} and it excels at handling them without
sacrificing the minimality requirement. Here is how one can express the task
for producing \cmd{release.tar} from the example discussed
in~\S\ref{sec-background-make}.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
"release.tar" %> \_ -> do
    need ["release.txt"]
    contents <- lines <$> readFile "release.txt"
    need contents
    system "tar" $ ["-cf", "result.tar"] ++ contents
\end{minted}
\vspace{1mm}

\noindent
We first declare the static dependency on \cmd{release.txt}, then read its
content and depend on it, dynamically. Finally, we specify the command line to
produce the resulting archive. Crucially, the archive will only be rebuilt if
one of the dependencies (static or dynamic) has changed.

\begin{figure}[h]
\centerline{\includegraphics[scale=0.28]{fig/make-example-cutoff.pdf}}
\vspace{-2mm}
\caption{An early cutoff example: modify the source \cmd{main.c} by adding a
comment. The rebuild is stopped after detecting that \cmd{main.o} is unchanged,
which indicates that \cmd{main.exe} does not need to be rebuilt.\label{fig-cutoff}}
\end{figure}

@@

\subsection{\Bazel: a cloud build system}
\label{sec-background-shake}

@@

\subsection{Abstract vocabulary for build systems}
\label{sec-background-vocabulary}
